01-introduction

### 2024-06-14 01-introduction.qmd
revisión párrafos chile

The Chilean case is particularly intriguing for studying market justice preferences. This country is characterized by acute and persistent economic inequality, which stands out in Latin America and among OECD countries [@flores_top_2020]. In Chile, the poorest 50% captures only 10% of the total income and has negative wealth, while the richest 1% receives almost 27% of the income and holds 49.6% of the wealth [@chancel_world_2022]. Much of this inequality is linked to the deep neoliberal reforms that institutionalized privatization and commodification of various economic sectors, including public services, introduced during the dictatorship (1973-1989) and expanded in democracy through concessions, demand credits, and specific regulatory frameworks [@ffrench-davis_reformas_2018]. This shift in economic policy has allowed the unprecedented emergence of markets in areas related to social reproduction, such as health, pensions, and education, with provision and access managed by private entities and segmented by individual payment capacity, heavily reliant on State subsidies [@boccardo_30_2020]. In health, although the majority of the population uses the public insurance system (78.9%), 15.3% are served by private insurers (CASEN, 2022). The pension system is based on individual capitalization, with mandatory contributions managed by private administrators investing in the financial market, currently involving 11 million contributors (Superintendencia de Pensiones, 2024). In education, 30.6% of school enrollment is in public schools, 54.0% in State-subsidized private schools, and 9.3% in fully private schools, generally attracting higher-income groups (Mineduc, 2023).

The provision and access to services in these areas under market principles correlate with the perceptions, preferences, and attitudes of Chilean society. Recent public opinion surveys conducted by the Center for Public Studies indicate that, generally, people prefer public insurance (64.1%) over private insurance (35.9%) if both cost the same. In pensions, Chileans show varied preferences regarding the public or private status of pension fund administrators; 52.3% favor both private and public institutions, 38.8% prefer only public institutions, and 8.6% prefer only private ones. This likely relates to the severe legitimacy crisis faced by private pension fund administrators, which has sparked social movements in recent years [@boccardo_30_2020]. Interestingly, in education, there is a clear preference for private schools over public ones if both cost the same and are equally distant, with 63% preferring private and 37% public.

>>>>

The Chilean case is particularly intriguing for studying market justice preferences. This country is characterized by acute and persistent economic inequality, which stands out in Latin America and among OECD countries [@flores_top_2020]. In Chile, the poorest 50% captures only 10% of the total income and has negative wealth, while the richest 1% receives almost 27% of the income and holds 49.6% of the wealth [@chancel_world_2022]. Much of this inequality has been attributed to the deep neoliberal reforms that institutionalized privatization and commodification of various economic sectors. Such reforms where introduced during the dictatorship (1973-1989) and expanded in democracy through concessions, demand credits, and specific regulatory frameworks [@ffrench-davis_reformas_2018]. This shift in economic policy has allowed the unprecedented emergence of markets in areas related to social reproduction, such as health, pensions, and education, with provision and access managed by private entities and segmented by individual payment capacity, heavily reliant on State subsidies [@boccardo_30_2020]. In health, although the majority of the population uses the public insurance system (78.9%), the pension system is based on individual capitalization, with mandatory contributions managed by private administrators investing in the financial market, currently involving 11 million contributors (Superintendencia de Pensiones, 2024). In education, 30.6% of school enrollment is in public schools, 54.0% in state-subsidized (voucher) private schools, and 9.3% in fully private schools, generally attracting higher-income groups (Mineduc, 2023).

Some surveys in Chile have considered the perception and support for the market provision in public areas. Recent public opinion surveys conducted by the Center for Public Studies (CEP) indicate that, generally, people prefer public insurance (64.1%) over private insurance (35.9%) if both cost the same. In pensions, Chileans show varied preferences regarding the public or private status of pension fund administrators; 52.3% favor both private and public institutions, 38.8% prefer only public institutions, and only 8.6% prefer only private ones. This likely relates to the severe legitimacy crisis faced by private pension fund administrators, which has sparked social movements in recent years [@boccardo_30_2020]. Interestingly, in education, there is a clear preference for private schools over public ones if both cost the same and are equally distant, with 63% preferring private and 37% public.
